---
title: "FINAL PROJECT"
author: "callie tansill-suddath"
date: "4/26/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown

#For my final project fot JOUR 772, I am looking at hospital preparedness during the COVID-19 pandemic in the United States. I am doing this by first comparing number of icu beds per county with county populations, then I will look at COVID death rate data per county.

#loading packages I might need and some I recall using from previous labs

```{r}
library(tidyverse)
library(dplyr)
library(janitor)
library(knitr)
library(lubridate)
library(tidycensus)
```

## loading in the data

```{r}
census_data <- read_csv("2018_county_census.csv")
hospital_data <- read_csv("hospital_data.csv")
```

#viewing census data

```{r}
glimpse(census_data)
view(census_data)
```

#viewing hospital data

```{r}
glimpse(hospital_data)
view(hospital_data)
```

```

STEP 1: 
Clean the hospital data
* group by statecountyfips
* summarise to add up all_beds
* that will leave us with a dataframe of two columns (statecountyfips, sum_all_beds)
```

```{r}
hospital_data %>%
group_by(statecountyfips)%>% 
summarise(all_beds = n())
```

STEP 2:
Get Census data 
Pull in from tidycensus, by county
* total population

Pull in from tidycensus by county
* Median household income

Join the two census dataframes on GEOID (which is the county fips code)

```{r}
census_api_key("d0ed7ea2cfbf5fd6b5e737d5eca2c50bd07155b3")

county_geodata <- get_acs(geography = "county",
              variables = "B01001_001", geometry = TRUE)

median_income <- get_acs(geography = "county", 
              variables = c(medincome = "B19013_001"), 
              year = 2018)

colnames(county_geodata)
county_geodata <- county_geodata %>% rename(statecountyfips = GEOID)


county_geodata_clean <- county_geodata %>%
  select(statecountyfips, NAME, estimate)

population_2018 <- census_data %>%
  clean_names()  %>%
  select(statecountyfips, geo_display,est2018_all)

colnames(population_2018)
```

```{R}
population_income <- county_geodata_clean %>%
  inner_join(population_2018, by=("statecountyfips")) %>%
  select(NAME, estimate, est2018_all)

colnames(population_income)
population_income <- population_income %>% rename(income = estimate)
population_income <- population_income %>% rename(population2018 = est2018_all)
```
STEP 3:
Load USDA ruralness spreadsheet

```{r}

#rural_data <- read_csv("ruralurbancodes2013.csv")

```


STEP 4:
Join census data to hospital data
* Join on GEOID = statecountyfips

Join this census_hospital sheet to ruralness sheet
* Join on GEOID

STEP 5:
* In this newsly joined sheet that has fips, beds, population, median household income and ruralness, use mutate to calculate beds per person by dividing beds by population. 

STEP 6
Start To do analysis we need
One data frame with the following features:
* Each row is a county
* One column each for the following:
  * Name of the county 
  * Name of the state 
  * fips code (five digts, two digit state, three digit county i.e. 01049)
  * Total beds and/or ICU Beds
  * Total population of the county
  * A computed column with beds per person or beds per 10,000 people or 100,000
  * Median Household income
  * A marker of ruralness for each county USDA thing
