---
title: "FINAL PROJECT"
author: "callie tansill-suddath"
date: "4/26/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown

#For my final project fot JOUR 772, I am looking at hospital preparedness during the COVID-19 pandemic in the United States. I am doing this by first comparing number of icu beds per county with county populations, then I will look at COVID death rate data per county.

#loading packages I might need and some I recall using from previous labs

```{r}
library(tidyverse)
library(dplyr)
library(janitor)
library(knitr)
library(lubridate)
```

## loading in the data

```{r}
census_data <- read_csv("2018_county_census.csv")
hospital_data <- read_csv("hospital_data.csv")
```

#viewing census data

```{r}
glimpse(census_data)
view(census_data)
```

#viewing hospital data

```{r}
glimpse(hospital_data)
view(hospital_data)
```

#cleaning so hospital data is lowercase

```{r}
hospital_data <- clean_names(hospital_data)
```

#mutate

```{r}
census_data <- census_data %>%
  mutate(geo_display = mdy(geo_display)) %>%
  select(geo_display, everything())


```
